Supp mat: time series paper

### 2017-07-07 writing/manuscript_draft.Rmd
changed code to allow relative directories

<!--
- intro long time series
- my general question
- primer on power
- more general power interpretation
- 2+ approaches
- what we do in this paper
-->

@@

```{r,echo=F,eval=F,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length. The horizontal, dashed line is our desired statistical power of 0.8. The vertical, dashed line is the minimum time required to achieve our desired statistical power. (c) Minimum time required for different values of the trend strength ($r$) and (d) minimum time required for different levels of population variability ($\\sigma$).\\label{fig:theoretical_approach}'}
setwd("~/Desktop/Research/time-series-project")
source("scripts/theoretical-models/calculate_power.R")
source("scripts/theoretical-models/simulate_pop.R")

@@

<!-- code to pull in databases-->
```{r, echo=F}
# Pull in original database from Keith et al. 2015 and the cleaned version after our series of filtering steps
  setwd("~/Desktop/Research/time-series-project")

#Load keith et al. database (from GPDD database)
  load('data/Keith_et_al_2015_data.Rdata')

#Load our cleaned data (see other code for cleaning steps)
  load('cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata') #Do i need to filter out non-significant populations as well (where no trend was present)
 # pop_info=pop_info[pop_info$trend_p_value<0.05,]
 # long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
 # long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
```

@@

```{r,echo=F}
# Pull in results of biological characteristics comparison to min time series length required
setwd("~/Desktop/Research/time-series-project")
load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```
For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics [@Myhrvold2015] like body size and generation length. All `r nrow(LPI_pop_info)` populations are within the Aves class. In the supplementary material, we explore how the minimum time required does not strongly correlate with any biological characteristics.

@@

```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance given power of 0.8.\\label{fig:min_time_dist}'}
# Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
 setwd("~/Desktop/Research/time-series-project")
 load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")

@@

```{r,echo=F,eval=F}
#summary(poisson_model)
```

<!--
# Potential figures
- Examples of how determining minimum time series works (probably in methods)
- What min time should scale with theoretically (trend, power, variability, etc) - another multi panel figure
- Mean min time by class or system
- Separate results for species, body size, location, biology. - multi panel figure

-->

@@

\begin{longtable}[]{@{}lrrrr@{}}
\caption{Output of generalized linear model to examine predictors of the
minimum time required for determing long-term population
trends.\label{table:model_output}}\tabularnewline
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endfirsthead
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endhead
(Intercept) & 3.7405513 & 0.0242742 & 154.09595 & 0\tabularnewline
abs\_overall\_trend & -104.0533454 & 3.0941097 & -33.62949 &
0\tabularnewline
autocorrelation & -0.4899533 & 0.0431769 & -11.34757 & 0\tabularnewline
variance & 18.7366110 & 0.6404458 & 29.25557 & 0\tabularnewline
\bottomrule
\end{longtable}

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Mimimum time required versus (a) generation length (years), (b)
litter size (n), (c) adult body mass (grams), (d) maximum longevity
(years), (e) egg mass (grams), and (f) incubation (days). The lines in
each plot represent the best fit line from linear
regression.\label{fig:biological_correlates}}
\end{figure}

>>>>

@@

```{r,echo=F,eval=F,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length. The horizontal, dashed line is our desired statistical power of 0.8. The vertical, dashed line is the minimum time required to achieve our desired statistical power. (c) Minimum time required for different values of the trend strength ($r$) and (d) minimum time required for different levels of population variability ($\\sigma$).\\label{fig:theoretical_approach}'}
source("../scripts/theoretical-models/calculate_power.R")
source("../scripts/theoretical-models/simulate_pop.R")

@@

<!-- code to pull in databases-->
```{r, echo=F}
# Pull in original database from Keith et al. 2015 and the cleaned version after our series of filtering steps

#Load keith et al. database (from GPDD database)
  load('../data/Keith_et_al_2015_data.Rdata')

#Load our cleaned data (see other code for cleaning steps)
  load('../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata') #Do i need to filter out non-significant populations as well (where no trend was present)
 # pop_info=pop_info[pop_info$trend_p_value<0.05,]
 # long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
 # long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
```

@@

```{r,echo=F}
# Pull in results of biological characteristics comparison to min time series length required
load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```
For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics [@Myhrvold2015] like body size and generation length. All `r nrow(LPI_pop_info)` populations are within the Aves class. In the supplementary material, we explore how the minimum time required does not strongly correlate with any biological characteristics.

@@

```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance given power of 0.8.\\label{fig:min_time_dist}'}
# Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
 load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")

@@

@@

### 2017-07-07 writing/supp_mat_draft.Rmd
changed code to allow relative directories

```{r,echo=F}
 setwd("~/Desktop/Research/time-series-project")
 load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
    pop = subset(long_dat,long_dat$ID=='7558')
    model_example = lm(pop$popvalue~c(1:nrow(pop)))
```

@@

<!-- example of how subsampling and empirical approach works -->
```{r, echo=F,message=F,warning=F,fig.cap='(a) Population size of Bigeye tuna (\\emph{Thunnus obesus}) over time. The line is the best fit line from linear regresssion. (b) Statistical power for different subsets of the time series in panel a.\\label{fig:empirical_approach_example}'}
# Example of subsampling empirical data
 setwd("~/Desktop/Research/time-series-project")
 load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
    pop = subset(long_dat,long_dat$ID=='7558') #13192 for Harp seals, 7558 for Thunnus obesus
    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
    mtext(text = '(a)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
    mtext(text='population size', side=2,line=4,cex=1.2)
# could insert all possible regressions of length n
source('scripts/calculate_power_metric.R')
    power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
    plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
    mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
# could also insert standard error bars
    abline(h=0.8,col='red',lwd=2,lty=2)
    abline(v=head(which(power_example>0.8),1),col='red',lwd=2,lty=2)

@@

```{r poisson_model, echo=F,include=T,fig.cap='Output of generalized linear model with a Poisson error structure for predicting the minimum time required with explanatory variables of the absolute value of the slope coefficient (or trend strength), temporal autocorrelation, and variability in population size.\\label{fig:poisson_model}'}
setwd("~/Desktop/Research/time-series-project")
load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")

@@

## Minimum time required and biological correlates
```{r, echo=F}
setwd("~/Desktop/Research/time-series-project")
load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```

@@

```{r, echo=F, fig.cap='Mimimum time required versus (a) generation length (years), (b) litter size (n), (c) adult body mass (grams), (d) maximum longevity (years), (e) egg mass (grams), and (f) incubation (days). The lines in each plot represent the best fit line from linear regression.\\label{fig:biological_correlates}',fig.height=6}
setwd("~/Desktop/Research/time-series-project")
load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
par(mfrow=c(3,2),oma=c(5,4,1,1),mar=c(5,3,0,0))

@@

# Pull in data
setwd("~/Desktop/Research/time-series-project")
load("~/Desktop/Research/time-series-project/cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
# Biological correlates plots
par(mfrow=c(2,2),oma=c(8,1,1.5,0.5),mar=c(0,4,0,1))

@@

```{r,echo=F,message=F,error=F,fig.cap='Minimum time required to assess long-term trends in abundance for values of statistical significance ($\\alpha$) and power ($1-\\beta$).\\label{fig:min_time_vs_alpha_beta}'}
# Bring in theoretical code
# Build plot of $T_min$ vs alpha and another versus beta
# Explain results after building figures
setwd("~/Desktop/Research/time-series-project")
source('scripts/calculate_slope.R') 
source('scripts/theoretical-models/simulate_pop.R')
# calculate_power = function(r,phi,sigma,trials,max.time,alpha){
#   trials   = trials
#   max.time = max.time
#   r        = r
#   phi      = phi
#   sigma    = sigma
#   
#   y=replicate(n=trials,simulate_pop(r,phi,sigma,max.time)) 
#   
#   power=sum((apply(y,2,calculate_p_value)<alpha) & sign(apply(y,2,calculate_slope))==sign(r))/trials 
#   
#   return(power)
# }
# 
# 
# min_time_needed= function(alpha,power){
#   years_to_sample=seq(2,30,by=1)
#   y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=0.5,sigma = 1,trials=100,alpha=alpha)
#   min_time_needed = years_to_sample[tail(which(y<power),1)] + 1
#   return(min_time_needed)
# }
# 
# alpha_values=seq(0.001,0.2,by=0.01)
# power_values=seq(0.65,0.95,by=0.01)
# parameter_combinations = expand.grid(alpha_values,power_values)
# names(parameter_combinations) = c('alpha','power')
# parameter_combinations$min_time_required = NA
# 
# for (i in 1:nrow(parameter_combinations)){
# 
#   parameter_combinations$min_time_required[i] =  min_time_needed(parameter_combinations[i,1],parameter_combinations[i,2])
#  
#   #print(i)
# }

@@

# Pull in data to prevent R from running for a long time to create data from scratch
parameter_combinations = read.csv('analysis-outputs/min_time_vs_alpha_and_power.csv',header=T)

@@

```{r,echo=F,message=F,error=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in log(abundance) given power of 0.8.\\label{fig:min_time_dist_log_pop}'}
setwd("~/Desktop/Research/time-series-project")
# load('cleaned-data/cleaned_timeseries_database.Rdata')
# 
# source('scripts/calculate_slope.R')
# source('scripts/calculate_power_metric.R')
# 
# for (j in 1:nrow(pop_info)){
#     pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
#     pop = pop[(nrow(pop)-34):nrow(pop),]
#     pop$popvalue=as.numeric(pop$popvalue)
#     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
#     ## Take the log of population size and add a small number
#     pop$popvalue = log(pop$popvalue + 0.001)
#     
#     pop_info$year[j] = pop$year[1]
#     pop_info$variance[j]= var(pop$popvalue)
#     pop_info$overall_trend[j] = calculate_slope(pop$popvalue)
#     pop_info$autocorrelation[j] = acf(pop$popvalue,plot = FALSE)$acf[2]
#     pop_info$trend_p_value[j]=calculate_p_value(pop$popvalue)
#     pop_info$popvalue = pop$popvalue[1] #initial abundance
#     pop_info$series_length = nrow(pop)
#     
# 
#      pop_info$min_time_for_power[j] = min_time_needed(pop$popvalue,0.05,0.8)
#     print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
#     
# }
# 
# pop_info=pop_info[pop_info$trend_p_value<0.05,]
# long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
# long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
# 
# save(pop_info,long_dat,long_time_names,file = 'cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression_for_log_pop.Rdata')

# LOAD in previously run script from this code chuck and make histogram plot
load('cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression_for_log_pop.Rdata')

@@

setwd("~/Desktop/Research/time-series-project")
#source('scripts/LemonShark_PopulationModel.R')
years_to_sample=seq(5,50,by=1)
#shark_power=sapply(years_to_sample,power)

load('analysis-outputs/example_power_for_shark_model.Rdata')
par(mfrow=c(1,1),oma=c(3,3,0.5,0.5),mar=c(2,2,0,0))
   plot(years_to_sample,shark_power,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
      abline(v=years_to_sample[tail(which(shark_power<0.8),1)+1],lty=2,lwd=2,col='red')
      abline(h=0.8,lty=2,lwd=2,col='red')
      mtext(text = 'time sampled' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3.5)
      #mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

#save(shark_power,file='analysis-outputs/example_power_for_shark_model.Rdata')

@@

```{r, echo=F, fig.cap='Example of calculating minimum time required for growth rate estimation. (a) European herring gull (\\emph{Larus argentatus}) scaled density over time, (b) mean and standard deviation of growth rate for subsamples of entire time series, and (c) the percent error between mean estimated growth rate and the true long-term growth rate. The vertical bar denotes the minimum time required to estimate growth rate within 20% error.\\label{fig:growth_rate}'}
#Example of using population growth rate instead of power
setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
source("scripts/calculate_geometric_growth.R")
load('cleaned-data/cleaned_timeseries_database.Rdata')

@@

  pop = subset(long_dat,long_dat$ID=='10008') #13192 for Harp seals, 10007 is California Gull, 10008 for European herring gull (Larus argentatus)
  source('scripts/create_subsamples.R')
  source('scripts/percent_error.R')
  aaa=calculate_average_geometric_growth(pop$popvalue)
  plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,las=1)
    #abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
    mtext(text = 'year',1,line=3,cex=1.4)
    mtext(text = 'population size',2,line=3.5,cex=1.4)
    mtext(text = '(a)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)

@@

```{r, echo=F,eval=T,fig.cap='Histogram of the minimum time required in order to estimate the long-term growth rate within 20% error.\\label{fig:min_time_growth_dist}'}
# Growth rates calculations for each species
setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
# load('cleaned-data/cleaned_timeseries_database.Rdata')
# 
# source('scripts/create_subsamples.R')
# source('scripts/calculate_geometric_growth.R')
# source('scripts/percent_error.R')
#  for (j in 1:nrow(pop_info)){
#     pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
#     pop$popvalue=as.numeric(pop$popvalue)
#     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
#     
#    # source('scripts/calculate_power_metric.R')
#     aaa=calculate_average_geometric_growth(pop$popvalue)
#     if (length(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.2))==0){
#       pop_info$min_time_growth_rate[j] = 2
#     }else{
#       pop_info$min_time_growth_rate[j] = tail(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.2),1)+1}
#     #print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
#  }
# save(pop_info,file = 'analysis-outputs/min_time_for_growth_rate0_2.Rdata')

load('analysis-outputs/min_time_for_growth_rate0_2.Rdata')

@@

```{r,echo=F,message=F,warning=F,fig.cap='(a) Time series for Bigeye tuna (\\emph{Thunnus obesus}) with corresponding fitted GAM model in red and (b) statistical power as a function of the number of years sampled. The horizontal ine at 0.8 indicates the minimum threshold for statistical power and the vertical line denotes the minimum time required to achieve 0.8 statistical power.\\label{fig:gam_example}'}
# show gam example side-by-side plot
 setwd("~/Desktop/Research/time-series-project")
 load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
 source('scripts/calculate_power_metric_gam.R')
  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
    pop = subset(long_dat,long_dat$ID=='7558') #13192 for Harp seals, 7558 for Thunnus obesus
    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
    example_gam_model=gam(pop$popvalue ~ s(pop$year,k=3),family=gaussian)
    points(pop$year,example_gam_model$fitted.values,col='red',type='l',lwd=2)
    mtext(text = '(a)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
    mtext(text='population size', side=2,line=4,cex=1.2)
# could insert all possible regressions of length n

@@

```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance according to a GAM model given statistical power of 0.8. The smoothing parameter was set to 3 for each population.\\label{fig:min_time_dist_gam}'}
# show gam results for minimium time required
setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
main_text = pop_info
# load('cleaned-data/cleaned_timeseries_database.Rdata')
#  source('scripts/calculate_power_metric_gam.R')
#  for (j in 1:nrow(pop_info)){
#     pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
#     pop$popvalue=as.numeric(pop$popvalue)
#     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
#   
#     pop$trend_p_value = calculate_p_value_gam(pop$popvalue)
#  
#     pop_info$min_time_for_power_gam[j] = min_time_needed(pop$popvalue,0.05,0.8)
#     print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
#  }
# 
# #pop_info$min_time_for_power_gam[pop_info$min_time_for_power_gam==-999]=33
# 
# pop_info=pop_info[pop_info$trend_p_value<0.05,]
# long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
# long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
# 
# save(pop_info,long_dat,long_time_names,file='analysis-outputs/min_time_required_GAM.Rdata')

load('analysis-outputs/min_time_required_GAM.Rdata')
# Create plot
par(mfrow=c(1,1))
min_time_gam = pop_info$min_time_for_power_gam[pop_info$min_time_for_power_gam!=-999]
plot(density(min_time_gam),las=1,ylab='',xlab='',cex.axis=1.2,main='',xlim=c(0,40))
mtext(text = 'Frequency',2,line=3,cex=1.2)
mtext(text = 'Minimum time required',1,line=3,cex=1.2)
```

>>>>

```{r,echo=F}
 #
 load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
    pop = subset(long_dat,long_dat$ID=='7558')
    model_example = lm(pop$popvalue~c(1:nrow(pop)))
```

@@

<!-- example of how subsampling and empirical approach works -->
```{r, echo=F,message=F,warning=F,fig.cap='(a) Population size of Bigeye tuna (\\emph{Thunnus obesus}) over time. The line is the best fit line from linear regresssion. (b) Statistical power for different subsets of the time series in panel a.\\label{fig:empirical_approach_example}'}
# Example of subsampling empirical data
 load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
    pop = subset(long_dat,long_dat$ID=='7558') #13192 for Harp seals, 7558 for Thunnus obesus
    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
    mtext(text = '(a)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
    mtext(text='population size', side=2,line=4,cex=1.2)
# could insert all possible regressions of length n
source("../scripts/calculate_power_metric.R",chdir=T)
    power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
    plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
    mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
# could also insert standard error bars
    abline(h=0.8,col='red',lwd=2,lty=2)
    abline(v=head(which(power_example>0.8),1),col='red',lwd=2,lty=2)

@@

```{r poisson_model, echo=F,include=T,fig.cap='Output of generalized linear model with a Poisson error structure for predicting the minimum time required with explanatory variables of the absolute value of the slope coefficient (or trend strength), temporal autocorrelation, and variability in population size.\\label{fig:poisson_model}'}

load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")

@@

## Minimum time required and biological correlates
```{r, echo=F}

load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```

@@

```{r, echo=F, fig.cap='Mimimum time required versus (a) generation length (years), (b) litter size (n), (c) adult body mass (grams), (d) maximum longevity (years), (e) egg mass (grams), and (f) incubation (days). The lines in each plot represent the best fit line from linear regression.\\label{fig:biological_correlates}',fig.height=6}

load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
par(mfrow=c(3,2),oma=c(5,4,1,1),mar=c(5,3,0,0))

@@

# Pull in data

load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
# Biological correlates plots
par(mfrow=c(2,2),oma=c(8,1,1.5,0.5),mar=c(0,4,0,1))

@@

```{r,echo=F,message=F,error=F,fig.cap='Minimum time required to assess long-term trends in abundance for values of statistical significance ($\\alpha$) and power ($1-\\beta$).\\label{fig:min_time_vs_alpha_beta}'}
# Bring in theoretical code
# Build plot of $T_min$ vs alpha and another versus beta
# Explain results after building figures

source('../scripts/calculate_slope.R') 
source('../scripts/theoretical-models/simulate_pop.R')
# calculate_power = function(r,phi,sigma,trials,max.time,alpha){
#   trials   = trials
#   max.time = max.time
#   r        = r
#   phi      = phi
#   sigma    = sigma
#   
#   y=replicate(n=trials,simulate_pop(r,phi,sigma,max.time)) 
#   
#   power=sum((apply(y,2,calculate_p_value)<alpha) & sign(apply(y,2,calculate_slope))==sign(r))/trials 
#   
#   return(power)
# }
# 
# 
# min_time_needed= function(alpha,power){
#   years_to_sample=seq(2,30,by=1)
#   y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=0.5,sigma = 1,trials=100,alpha=alpha)
#   min_time_needed = years_to_sample[tail(which(y<power),1)] + 1
#   return(min_time_needed)
# }
# 
# alpha_values=seq(0.001,0.2,by=0.01)
# power_values=seq(0.65,0.95,by=0.01)
# parameter_combinations = expand.grid(alpha_values,power_values)
# names(parameter_combinations) = c('alpha','power')
# parameter_combinations$min_time_required = NA
# 
# for (i in 1:nrow(parameter_combinations)){
# 
#   parameter_combinations$min_time_required[i] =  min_time_needed(parameter_combinations[i,1],parameter_combinations[i,2])
#  
#   #print(i)
# }

@@

# Pull in data to prevent R from running for a long time to create data from scratch
parameter_combinations = read.csv('../analysis-outputs/min_time_vs_alpha_and_power.csv',header=T)

@@

```{r,echo=F,message=F,error=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in log(abundance) given power of 0.8.\\label{fig:min_time_dist_log_pop}'}

# load('../cleaned-data/cleaned_timeseries_database.Rdata')
# 
# source('../scripts/calculate_slope.R')
# source('../scripts/calculate_power_metric.R')
# 
# for (j in 1:nrow(pop_info)){
#     pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
#     pop = pop[(nrow(pop)-34):nrow(pop),]
#     pop$popvalue=as.numeric(pop$popvalue)
#     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
#     ## Take the log of population size and add a small number
#     pop$popvalue = log(pop$popvalue + 0.001)
#     
#     pop_info$year[j] = pop$year[1]
#     pop_info$variance[j]= var(pop$popvalue)
#     pop_info$overall_trend[j] = calculate_slope(pop$popvalue)
#     pop_info$autocorrelation[j] = acf(pop$popvalue,plot = FALSE)$acf[2]
#     pop_info$trend_p_value[j]=calculate_p_value(pop$popvalue)
#     pop_info$popvalue = pop$popvalue[1] #initial abundance
#     pop_info$series_length = nrow(pop)
#     
# 
#      pop_info$min_time_for_power[j] = min_time_needed(pop$popvalue,0.05,0.8)
#     print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
#     
# }
# 
# pop_info=pop_info[pop_info$trend_p_value<0.05,]
# long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
# long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
# 
# save(pop_info,long_dat,long_time_names,file = '../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression_for_log_pop.Rdata')

# LOAD in previously run script from this code chuck and make histogram plot
load('../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression_for_log_pop.Rdata')

@@

#source('../scripts/LemonShark_PopulationModel.R')
years_to_sample=seq(5,50,by=1)
#shark_power=sapply(years_to_sample,power)

load('../analysis-outputs/example_power_for_shark_model.Rdata')
par(mfrow=c(1,1),oma=c(3,3,0.5,0.5),mar=c(2,2,0,0))
   plot(years_to_sample,shark_power,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
      abline(v=years_to_sample[tail(which(shark_power<0.8),1)+1],lty=2,lwd=2,col='red')
      abline(h=0.8,lty=2,lwd=2,col='red')
      mtext(text = 'time sampled' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3.5)
      #mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

#save(shark_power,file='../analysis-outputs/example_power_for_shark_model.Rdata')

@@

```{r, echo=F, fig.cap='Example of calculating minimum time required for growth rate estimation. (a) European herring gull (\\emph{Larus argentatus}) scaled density over time, (b) mean and standard deviation of growth rate for subsamples of entire time series, and (c) the percent error between mean estimated growth rate and the true long-term growth rate. The vertical bar denotes the minimum time required to estimate growth rate within 20% error.\\label{fig:growth_rate}'}
#Example of using population growth rate instead of power
source("../scripts/calculate_geometric_growth.R")
load('../cleaned-data/cleaned_timeseries_database.Rdata')

@@

  pop = subset(long_dat,long_dat$ID=='10008') #13192 for Harp seals, 10007 is California Gull, 10008 for European herring gull (Larus argentatus)
  source('../scripts/create_subsamples.R')
  source('../scripts/percent_error.R')
  aaa=calculate_average_geometric_growth(pop$popvalue)
  plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,las=1)
    #abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
    mtext(text = 'year',1,line=3,cex=1.4)
    mtext(text = 'population size',2,line=3.5,cex=1.4)
    mtext(text = '(a)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)

@@

```{r, echo=F,eval=T,fig.cap='Histogram of the minimum time required in order to estimate the long-term growth rate within 20% error.\\label{fig:min_time_growth_dist}'}
# Growth rates calculations for each species
# load('../cleaned-data/cleaned_timeseries_database.Rdata')
# 
# source('../scripts/create_subsamples.R')
# source('../scripts/calculate_geometric_growth.R')
# source('../scripts/percent_error.R')
#  for (j in 1:nrow(pop_info)){
#     pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
#     pop$popvalue=as.numeric(pop$popvalue)
#     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
#     
#    # source('../scripts/calculate_power_metric.R')
#     aaa=calculate_average_geometric_growth(pop$popvalue)
#     if (length(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.2))==0){
#       pop_info$min_time_growth_rate[j] = 2
#     }else{
#       pop_info$min_time_growth_rate[j] = tail(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.2),1)+1}
#     #print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
#  }
# save(pop_info,file = '../analysis-outputs/min_time_for_growth_rate0_2.Rdata')

load('../analysis-outputs/min_time_for_growth_rate0_2.Rdata')

@@

```{r,echo=F,message=F,warning=F,fig.cap='(a) Time series for Bigeye tuna (\\emph{Thunnus obesus}) with corresponding fitted GAM model in red and (b) statistical power as a function of the number of years sampled. The horizontal ine at 0.8 indicates the minimum threshold for statistical power and the vertical line denotes the minimum time required to achieve 0.8 statistical power.\\label{fig:gam_example}'}
# show gam example side-by-side plot

 load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
 source('../scripts/calculate_power_metric_gam.R')
  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
    pop = subset(long_dat,long_dat$ID=='7558') #13192 for Harp seals, 7558 for Thunnus obesus
    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
    example_gam_model=gam(pop$popvalue ~ s(pop$year,k=3),family=gaussian)
    points(pop$year,example_gam_model$fitted.values,col='red',type='l',lwd=2)
    mtext(text = '(a)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
    mtext(text='population size', side=2,line=4,cex=1.2)
# could insert all possible regressions of length n

@@

```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance according to a GAM model given statistical power of 0.8. The smoothing parameter was set to 3 for each population.\\label{fig:min_time_dist_gam}'}
# show gam results for minimium time required
load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
main_text = pop_info
# load('../cleaned-data/cleaned_timeseries_database.Rdata')
#  source('../scripts/calculate_power_metric_gam.R')
#  for (j in 1:nrow(pop_info)){
#     pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
#     pop$popvalue=as.numeric(pop$popvalue)
#     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
#   
#     pop$trend_p_value = calculate_p_value_gam(pop$popvalue)
#  
#     pop_info$min_time_for_power_gam[j] = min_time_needed(pop$popvalue,0.05,0.8)
#     print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
#  }
# 
# #pop_info$min_time_for_power_gam[pop_info$min_time_for_power_gam==-999]=33
# 
# pop_info=pop_info[pop_info$trend_p_value<0.05,]
# long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
# long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
# 
# save(pop_info,long_dat,long_time_names,file='../analysis-outputs/min_time_required_GAM.Rdata')

load('../analysis-outputs/min_time_required_GAM.Rdata')
# Create plot
par(mfrow=c(1,1))
min_time_gam = pop_info$min_time_for_power_gam[pop_info$min_time_for_power_gam!=-999]
plot(density(min_time_gam),las=1,ylab='',xlab='',cex.axis=1.2,main='',xlim=c(0,40))
mtext(text = 'Frequency',2,line=3,cex=1.2)
mtext(text = 'Minimum time required',1,line=3,cex=1.2)
```
